Computing exact restricted isometry constants via mixed integer semidefinite programming

### 2014-12-18 RIP-MISDP-Paper/Paper_v1.tex
try \ones and replace ^T by \T

In this section we will discuss two known SDP relaxations to compute an upper bound on the restricted isometry constant $\delta_k$ in (\ref{rip}), as they have some ideas in common with the mixed integer semidefinite program we want to 
propose and will also be checked against in the section on numerical results. The idea of these two relaxations as well as the MISDP in the next section is that the optimal constant $\alpha_k$ in (\ref{asymRIP}) can be computed
with the non-convex quadratic optimization problem
\begin{align}\label{QP}
 \text{min} \quad & \|Ax\|_2^2 \nonumber \\
 \text{s.t.} \quad & \|x\|_2^2 = 1 \tag{QP} \\
 & \|x\|_0 \leq k. \nonumber
\end{align}
The corresponding maximization problem can be solved to compute $\beta_k$. In \cite{Asp07} the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ is tackled by the technique of semidefinite lifting first introduced in \cite{GW95} 
for the max-cut problem. The idea is to use a new matrix variable $X=xx^T$ for the quadratic terms, where the condition $X=xx^T$ is then enforced by the equivalent constraints $X \succeq 0$ and $\textbf{Rk}(X) = 1$, where 
$X \succeq 0$ means that $X$ is symmetric and positive semidefinite. In the next step the non-convex rank constraint is then relaxed and the inequality $\|x\|_0 \leq k$ is substituted by the coarser constraint $1^T|X|1 \leq k$ to form
the relaxation
\begin{align}\label{Asp07}
 \text{min} \quad & \textbf{Tr}(A^TAX) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \nonumber \\
 & 1^T|X|1 \leq k \tag{Asp07} \\
 & X \succeq 0. \nonumber
\end{align}

@@

\begin{equation}\label{phi}
 \phi(\rho) = \max_{\|x\|_2 \leq 1} x^T \Sigma x - \rho \|x\|_0
\end{equation}
is tackled for a symmetric positive semidefinite matrix $\Sigma$, which for $\Sigma = A^TA$, by duality, gives an upper bound on $\beta_k$ via

\begin{equation*}
 \beta_k^2 \leq \inf_{\rho \geq 0} \phi(\rho) + \rho k.
\end{equation*}
To solve (\ref{phi}) again the semidefinite lifting technique is used to compute an upper bound on $\phi$ via
\begin{align}\label{Asp08}
 \text{max} \quad & \sum_{i=1}^n\textbf{Tr}(P_iB_i) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \nonumber \\
 & X \succeq 0 \tag{Asp08} \\
 & X \succeq P_i \succeq 0 \quad \forall i \leq n, \nonumber
\end{align}
for $B_i = b_ib_i^T - \rho I,$ where $b_i$ is the $i$-th column of a squareroot of $\Sigma$. To compute the parameter $\rho$, the authors of \cite{Asp07} propose to minimize over the interval
\begin{equation*}
 \max_{i \notin I} (b_i^Tx)^2 \leq \rho \leq  \min_{i \in I} (b_i^Tx)^2
\end{equation*}
the convex function
\begin{equation*}
 \text{gap}(\rho) = \lambda_{\max} \left( \sum_{i=1}^n Y_i \right) - \sum_{i \in I} ((b_i^Tx)^2 - \rho),
\end{equation*}
where $I$ is a given sparsity pattern, found by some heuristic for example, $x$ the largest eigenvector of $\sum_{i \in I} b_i b_i^T$ and
\begin{equation*}
 Y_i = \begin{cases}\frac{B_ixx^TB_i}{x^TB_ix}, & \text{for } i \in I \\ \max\{0, \rho \frac{b_i^Tb_i - \rho}{\rho - (b_i^Tx)^2} \} \frac{(I-xx^T)b_ib_i^T(I-xx^T)}{\|(I-xx^T)b_i\|_2^2}, & \text{for } i \in I^c\end{cases}. 
 %x in the second case sill as above?
\end{equation*}

@@

For the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ we will again use semidefinite lifting as $\|Ax\|_2^2 = x^TA^TAx = \textbf{Tr}(A^TAX)$ and $\textbf{Tr}(X) = \sum_{i=1}^n x_i^2$. 
The cardinality-constraint in (\ref{QP}) can equivalently be written using binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i \leq k$. We can thus rewrite (\ref{QP}) equivalently to
% In this section we want to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}) using mixed integer semidefinite programming. A non-convex quadratic optimization problem to compute $\alpha_k$ can be formulated as
% 
% \begin{align}\label{QP}
%  \text{min} \quad & \|Ax\|_2^2 \nonumber \\
%  \text{s.t.} \quad & \|x\|_2^2 = 1 \tag{QP} \\
%  & \|x\|_0 \leq k. \nonumber
% \end{align}
% 
% By switching $\text{min}$ for $\text{max}$ in (\ref{QP}) we also get a corresponding optimization problem to compute $\beta_k$ and therefore also $\gamma_k$ and $\delta_k$.
% The cardinality-constraint in (\ref{QP}) can equivalently be written using binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i \leq k$. For the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ we will
% use the technique of semidefinite lifting first introduced in \cite{GW95} for the max-cut problem. The idea is to use a new matrix variable $X=xx^T$ for the quadratic terms, where the condition $X=xx^T$ is then enforced by the 
% equivalent constraints $X \succeq 0$ and $\textbf{Rk}(X) = 1$, where $X \succeq 0$ means that $X$ is symmetric and positive semidefinite. Using that $\|Ax\|_2^2 = x^TA^TAx = \textbf{Tr}(A^TAX)$ and 
% $\textbf{Tr}(X) = \sum_{i=1}^n x_i^2$, we can thus rewrite (\ref{QP}) equivalently to
\begin{align}\label{Rk1MISDP}
 \text{min} \quad & \textbf{Tr}(A^TAX) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \forall j \leq n \nonumber \\
 & \sum_{i=1}^n z_i \leq k \tag{Rk1MISDP} \\
 & \textbf{Rk}(X) = 1 \nonumber \\
 & X \succeq 0  \nonumber \\
 & z \in \{0,1\}^n \nonumber
\end{align}

The next step would then be to relax the non-convex rank constraint again to arrive at the following relaxation which computes a lower bound on the optimal constant $\alpha_k$ in (\ref{asymRIP}):
\begin{align}\label{MISDP}
 \text{min} \quad & \textbf{Tr}(A^TAX) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \forall j \leq n \nonumber \\
 & \sum_{i=1}^n z_i \leq k \tag{MISDP} \\
 & X \succeq 0  \nonumber \\
 & z \in \{0,1\}^n \nonumber
\end{align}

@@

\begin{thm}\label{Rk1thm}
 For every $A \in \R^{m \times n}$ and $k \geq 1$ there exists an optimal solution $X^*, z^*$ to (\ref{MISDP}) such that $\textbf{Rk}(X^*) = 1$. 
\end{thm}
\begin{proof}
 Let $\hat{X}, \hat{z}$ be an optimal solution to (\ref{MISDP}) (because of $\textbf{Tr}(X) = 1$ the problem is bounded and $X_{11} = 1, X_{ij} = 0 \ \forall i,j \neq 1, z_1 = 1, z_i = 0 \ \forall i \neq 0$ is always feasible)
 with $\sum_{i=1}^n z_i = \ell \leq k$. Define
 \begin{align}\label{Proj}
 \text{min} \quad & \textbf{Tr}(\tilde{A}^T\tilde{A}X) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \tag{Proj} \\
 & X \succeq 0 , X \in S_\ell, \nonumber
 \end{align}
where $\tilde{A} = (A_{i_1}, A_{i_2}, ..., A_{i_\ell})$ consists of those columns $\{{i_1}, {i_2}, ..., {i_\ell}\} =: T$ of $A$ for which $\hat{z}_{i_k} = 1$. \\
Now let $\tilde{X}$ be the matrix consisting of those rows and columns 
$i_k$ of $\hat{X}$ such that $\hat{z}_{i_k} = 1$. Then $\hat{X}$ is feasible for (\ref{Proj}) with the same objective value as the optimal solution $\hat{X}$ of (\ref{MISDP}), as we only removed zero rows and columns of $\hat{X}$ 
and $(A^TA)\hat{X}$, which do not influence the trace or the positive semidefiniteness. \\
On the other hand we can lift any optimal solution $\tilde{X}$ of (\ref{Proj}) to a feasible point for (\ref{MISDP}) with identical objective value by setting $\bar{z}_i$ to one if and only if $i \in T$
and $\bar{X}_{i_j i_k} = \tilde{X}_{jk}$ if and only if $i_j, i_k \in T$ and $\bar{X}_{ij}=0$ otherwise. \\
So the optimal objective values of (\ref{Proj}) and (\ref{MISDP}) agree and with this construction we also get that 
$\textbf{Rk}(\bar{X}) = \textbf{Rk}(\tilde{X})$, so it suffices to show that (\ref{Proj}) always allows for a solution of rank one.
Let $\breve{X}$ be an extreme point of (\ref{Proj}) (as the set of optimal points of (\ref{Proj}) is a non-empty, convex, compact subset of $\R^{\ell(\ell+1)/2}$ such a point exists as a consequence of corollary 18.5.1 in
\cite{Roc70}) with corresponding face $F=\{\breve{X}\}$. Then using theorem \ref{patakiLemma} on (\ref{Proj}) with $\tilde{m} = 1$ LP-constraint gives
\begin{equation*}
 \frac{1}{2}\textbf{Rk}(\breve{X})\cdot(\textbf{Rk}(\breve{X})+1) \leq \tilde{m} + \textbf{dim}(F) = 1 + 0 = 1.
\end{equation*}
As $\breve{X}=0$ is infeasible for $\textbf{Tr}(X)=1$, we must have $\textbf{Rk}(\breve{X}) = 1$, which can then be lifted to an optimal solution of (\ref{MISDP}) with rank one.
\end{proof}

>>>>

In this section we will discuss two known SDP relaxations to compute an upper bound on the restricted isometry constant $\delta_k$ in (\ref{rip}), as they have some ideas in common with the mixed integer semidefinite program we want to 
propose and will also be checked against in the section on numerical results. The idea of these two relaxations as well as the MISDP in the next section is that the optimal constant $\alpha_k$ in (\ref{asymRIP}) can be computed
with the non-convex quadratic optimization problem
\begin{align}\label{QP}
 \text{min} \quad & \|Ax\|_2^2 \nonumber \\
 \text{s.t.} \quad & \|x\|_2^2 = 1 \tag{QP} \\
 & \|x\|_0 \leq k. \nonumber
\end{align}
The corresponding maximization problem can be solved to compute $\beta_k$. In \cite{Asp07} the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ is tackled by the technique of semidefinite lifting first introduced in \cite{GW95} 
for the max-cut problem. The idea is to use a new matrix variable $X=xx\T$ for the quadratic terms, where the condition $X=xx\T$ is then enforced by the equivalent constraints $X \succeq 0$ and $\textbf{Rk}(X) = 1$, where 
$X \succeq 0$ means that $X$ is symmetric and positive semidefinite. In the next step the non-convex rank constraint is then relaxed and the inequality $\|x\|_0 \leq k$ is substituted by the coarser constraint $1\T|X|1 \leq k$ to form
the relaxation
\begin{align}\label{Asp07}
 \text{min} \quad & \textbf{Tr}(A\T A X) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \nonumber \\
 & \ones\T|X|\ones \leq k \tag{Asp07} \\
 & X \succeq 0. \nonumber
\end{align}

@@

\begin{equation}\label{phi}
 \phi(\rho) = \max_{\|x\|_2 \leq 1} x\T \Sigma x - \rho \|x\|_0
\end{equation}
is tackled for a symmetric positive semidefinite matrix $\Sigma$, which for
$\Sigma = A\T A$, by duality, gives an upper bound on $\beta_k$ via

\begin{equation*}
 \beta_k^2 \leq \inf_{\rho \geq 0} \phi(\rho) + \rho k.
\end{equation*}
To solve (\ref{phi}) again the semidefinite lifting technique is used to compute an upper bound on $\phi$ via
\begin{align}\label{Asp08}
 \text{max} \quad & \sum_{i=1}^n\textbf{Tr}(P_iB_i) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \nonumber \\
 & X \succeq 0 \tag{Asp08} \\
 & X \succeq P_i \succeq 0 \quad \forall i \leq n, \nonumber
\end{align}
for $B_i = b_ib_i\T - \rho I,$ where $b_i$ is the $i$-th column of a squareroot of $\Sigma$. To compute the parameter $\rho$, the authors of \cite{Asp07} propose to minimize over the interval
\begin{equation*}
 \max_{i \notin I} (b_i\T x)^2 \leq \rho \leq  \min_{i \in I} (b_i\T x)^2
\end{equation*}
the convex function
\begin{equation*}
 \text{gap}(\rho) = \lambda_{\max} \left( \sum_{i=1}^n Y_i \right) -
 \sum_{i \in I} ((b_i\T x)^2 - \rho),
\end{equation*}
where $I$ is a given sparsity pattern, found by some heuristic for example, $x$ the largest eigenvector of $\sum_{i \in I} b_i b_i\T$ and
\begin{equation*}
 Y_i = \begin{cases}\frac{B_ixx\T B_i}{x\T B_ix}, & \text{for } i \in I \\
   \max\{0, \rho \frac{b_i\T b_i - \rho}{\rho - (b_i\T x)^2} \} \frac{(I-xx\T)b_ib_i\T(I-xx\T)}{\|(I-xx\T)b_i\|_2^2}, & \text{for } i \in I^c\end{cases}. 
 %x in the second case sill as above?
\end{equation*}

@@

For the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ we will
again use semidefinite lifting as $\|Ax\|_2^2 = x\T A\T Ax =
\textbf{Tr}(A\T A X)$ and $\textbf{Tr}(X) = \sum_{i=1}^n x_i^2$. 
The cardinality-constraint in (\ref{QP}) can equivalently be written using binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i \leq k$. We can thus rewrite (\ref{QP}) equivalently to
% In this section we want to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}) using mixed integer semidefinite programming. A non-convex quadratic optimization problem to compute $\alpha_k$ can be formulated as
% 
% \begin{align}\label{QP}
%  \text{min} \quad & \|Ax\|_2^2 \nonumber \\
%  \text{s.t.} \quad & \|x\|_2^2 = 1 \tag{QP} \\
%  & \|x\|_0 \leq k. \nonumber
% \end{align}
% 
% By switching $\text{min}$ for $\text{max}$ in (\ref{QP}) we also get a corresponding optimization problem to compute $\beta_k$ and therefore also $\gamma_k$ and $\delta_k$.
% The cardinality-constraint in (\ref{QP}) can equivalently be written using binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i \leq k$. For the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ we will
% use the technique of semidefinite lifting first introduced in \cite{GW95} for the max-cut problem. The idea is to use a new matrix variable $X=xx\T$ for the quadratic terms, where the condition $X=xx\T$ is then enforced by the 
% equivalent constraints $X \succeq 0$ and $\textbf{Rk}(X) = 1$, where $X
% \succeq 0$ means that $X$ is symmetric and positive semidefinite. Using
% that $\|Ax\|_2^2 = x\T A\T Ax = \textbf{Tr}(A\T A X)$ and 
% $\textbf{Tr}(X) = \sum_{i=1}^n x_i^2$, we can thus rewrite (\ref{QP}) equivalently to
\begin{align}\label{Rk1MISDP}
 \text{min} \quad & \textbf{Tr}(A\T A X) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \forall j \leq n \nonumber \\
 & \sum_{i=1}^n z_i \leq k \tag{Rk1MISDP} \\
 & \textbf{Rk}(X) = 1 \nonumber \\
 & X \succeq 0  \nonumber \\
 & z \in \{0,1\}^n \nonumber
\end{align}

The next step would then be to relax the non-convex rank constraint again to arrive at the following relaxation which computes a lower bound on the optimal constant $\alpha_k$ in (\ref{asymRIP}):
\begin{align}\label{MISDP}
 \text{min} \quad & \textbf{Tr}(A\T A X) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \forall j \leq n \nonumber \\
 & \sum_{i=1}^n z_i \leq k \tag{MISDP} \\
 & X \succeq 0  \nonumber \\
 & z \in \{0,1\}^n \nonumber
\end{align}

@@

\begin{thm}\label{Rk1thm}
 For every $A \in \R^{m \times n}$ and $k \geq 1$ there exists an optimal solution $X^*, z^*$ to (\ref{MISDP}) such that $\textbf{Rk}(X^*) = 1$. 
\end{thm}
\begin{proof}
 Let $\hat{X}, \hat{z}$ be an optimal solution to (\ref{MISDP}) (because of $\textbf{Tr}(X) = 1$ the problem is bounded and $X_{11} = 1, X_{ij} = 0 \ \forall i,j \neq 1, z_1 = 1, z_i = 0 \ \forall i \neq 0$ is always feasible)
 with $\sum_{i=1}^n z_i = \ell \leq k$. Define
 \begin{align}\label{Proj}
 \text{min} \quad & \textbf{Tr}(\tilde{A}\T\tilde{A}X) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \tag{Proj} \\
 & X \succeq 0 , X \in S_\ell, \nonumber
 \end{align}
where $\tilde{A} = (A_{i_1}, A_{i_2}, ..., A_{i_\ell})$ consists of those columns $\{{i_1}, {i_2}, ..., {i_\ell}\} =: T$ of $A$ for which $\hat{z}_{i_k} = 1$. \\
Now let $\tilde{X}$ be the matrix consisting of those rows and columns 
$i_k$ of $\hat{X}$ such that $\hat{z}_{i_k} = 1$. Then $\hat{X}$ is feasible for (\ref{Proj}) with the same objective value as the optimal solution $\hat{X}$ of (\ref{MISDP}), as we only removed zero rows and columns of $\hat{X}$ 
and $(A\T A)\hat{X}$, which do not influence the trace or the positive semidefiniteness. \\
On the other hand we can lift any optimal solution $\tilde{X}$ of (\ref{Proj}) to a feasible point for (\ref{MISDP}) with identical objective value by setting $\bar{z}_i$ to one if and only if $i \in T$
and $\bar{X}_{i_j i_k} = \tilde{X}_{jk}$ if and only if $i_j, i_k \in T$ and $\bar{X}_{ij}=0$ otherwise. \\
So the optimal objective values of (\ref{Proj}) and (\ref{MISDP}) agree and with this construction we also get that 
$\textbf{Rk}(\bar{X}) = \textbf{Rk}(\tilde{X})$, so it suffices to show that (\ref{Proj}) always allows for a solution of rank one.
Let $\breve{X}$ be an extreme point of (\ref{Proj}) (as the set of optimal points of (\ref{Proj}) is a non-empty, convex, compact subset of $\R^{\ell(\ell+1)/2}$ such a point exists as a consequence of corollary 18.5.1 in
\cite{Roc70}) with corresponding face $F=\{\breve{X}\}$. Then using theorem \ref{patakiLemma} on (\ref{Proj}) with $\tilde{m} = 1$ LP-constraint gives
\begin{equation*}
 \frac{1}{2}\textbf{Rk}(\breve{X})\cdot(\textbf{Rk}(\breve{X})+1) \leq \tilde{m} + \textbf{dim}(F) = 1 + 0 = 1.
\end{equation*}
As $\breve{X}=0$ is infeasible for $\textbf{Tr}(X)=1$, we must have $\textbf{Rk}(\breve{X}) = 1$, which can then be lifted to an optimal solution of (\ref{MISDP}) with rank one.
\end{proof}
